---
title: "MAPITR Introduction -- Simulated Data"
author: "Michael C Turchin"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{MAPITR Introduction -- Simulated Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This is a simple introductory vignette to the `MAPITR` package. In this introductory example, we will use a simulated dataset to run `MAPITR` and explore some basic output. The dataset is provided with the package.

## Running `MAPITR`

For this introductory vignette, we will be using `MAPITR_SimData_Phenotype`, `MAPITR_SimData_Genotypes`, and `MAPITR_SimData_Pathways`. These are simulated datasets created for the purposes of this vignette. To load 'MAPITR' and the datasets, run the following:

```{r}
library("MAPITR");

data(MAPITR_SimData_Phenotype, MAPITR_SimData_Genotypes, MAPITR_SimData_Pathways)
```

First, we'll take a look at the formats of our simulated input data:

```{r}
MAPITR_SimData_Genotypes[1:10,1:10]
```

Here you should see the first ten rows and columns of our genotype matrix. The genotype matrix is n x p, so the rows are our n individuals and the columns are our p SNPs. SNP genotypes, the entries of each column, are assumed to be 0, 1, or 2 -- so note, diploid information is assumed to have been collaspsed into a single value for each SNP. For example, the typical diploid setup of 0/0, 0/1, and 1/1, representing either reference allele (0) or alternative allele (1) for both chromosomal copies in an individual, is now instead represented as 0 (0/0), 1 (0/1), or 2 (1/1).

Also note that there can be no genotype missingness present in the genotype matrix. This is necessary due to how the genetic relatedness matrices (GRMs), required components for the MAPITR variance component model, are constructed. For each entry in the genotype matrix, the only acceptable values are 0, 1, and 2. 

```{r}
head(MAPITR_SimData_Phenotype)
```

Here you should see the first ten values of our input phenotype. `MAPITR` only runs on one phenotype at a time currently, so the main function only requires a single vector of phenotypic values as input. 


```{r}
head(MAPITR_SimData_Pathways)
```

Here you should see a matrix with two columns. The first column should list the name of each pathway being analyzed. The second column should be a list of comma-separated column indices, representing the set of SNPs belonging to each associated pathway. These indices will be used by `MAPITR` to extract the proper SNPs per pathway from the input genotype matrix to conduct the marginal epistasis analysis. Note that the values for this second column are not the SNP IDs (eg rsID#s) or column headers (if your genotype matrix has them), they are the specific column locations in the input genotype matrix for each SNP.

Now to run `MAPITR`, we use the following code:

```{r}
MAPITR_Output <- MAPITR(MAPITR_SimData_Genotypes, MAPITR_SimData_Phenotype, MAPITR_SimData_Pathways)
```

And to see the output `MAPITR()` provides, we can run the following:

```{r}
head(MAPITR_Output$Results)
```

The output shown from `MAPITR_Output$Results` should be a matrix with four columns. The first column should be the list of pathway names being analyzed (same order as input file), the second column should be the `MAPITR` p-values for each pathway, the third column should be the estimate of the epistatic GRM variance component ('Q' in the [model][MAPITR-paper]) for each pathway, and the fourth column should be the proportion of variance explained (PVE) for the phenotype by each pathway. Note the PVE estimates here are for pathways that generally have overlapping variants included, so the interpretation has a different meaning than is typically thought of under single variant models where PVE can theoretically sum to 1 across all variants. A priori it is unclear what PVE should theoretically sum to since variant overlap across pathways will vary depending on the list of pathways used. 

In this example, there should be 1 pathway that has a p-value below the Bonferonni-corrected threshold of .01 (.05 / 5, the number of pathways tested):

```{r}
MAPITR_Output$Results[MAPITR_Output$Results[,2] < .01,]
```

For further details on the `MAPITR` model, please see the associated [manuscript][MAPITR-paper]. 

[MAPITR-paper]: http://biorxiv.org/
